Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-03 source/sec_overparameterized.tex
updated overparameterized

\subsection{Overparameterized Quadratic Problems}
We now turn to quadratic problems
and demonstrate how the general bound in Theorem~\ref{thm:overparameterized_general}
can be sharpened to explicitly depend on the condition number of the
system, and how this rate can be further improved with a new FedAvg
algorithm with momentum updates.

The local device objectives are now given by the sum of squares {\small$F_{k}(\mathbf{w})=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(\mathbf{w}^{T}\mathbf{x}_{k}^{j}-z_{k}^{j})^{2}$},
and there exists $\mathbf{w}^{\ast}$ such that $F(\mathbf{w}^{\ast})\equiv0$.
Define the local Hessian matrix as $H^{k}:=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}$, and the stochastic Hessian matrix as $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$. 
\begin{comment}
In general $H$ has zero eigevalues. However, because the null space
of $H$ and range of $H$ are orthogonal, in our subsequence analysis
it suffices to project $\overline{\mathbf{w}}_{t}-\mathbf{w}^{\ast}$
onto the range of $H$, thus we may restrict to the non-zero eigenvalue
of $H$. We can use $\mathbf{w}^{\ast T}\mathbf{x}_{k}^{j}-z_{k}^{j}\equiv0$
to rewrite the local objectives as $F_{k}(\mathbf{w})=\frac{1}{2}\langle\mathbf{w}-\mathbf{w}^{\ast},H^{k}(\mathbf{w}-\mathbf{w}^{\ast})\rangle\equiv\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H^{k}}^{2}$
so that $F(\mathbf{w})=\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H}^{2}$.
\end{comment}
\begin{comment}
\begin{align*}
F_{k}(w) & =\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(w^{T}x_{k,j}-z_{k,j}-(w^{\ast T}x_{k,j}-z_{k,j}))^{2}=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}((w-w^{\ast})^{T}x_{k,j})^{2}\\
& =\frac{1}{2}\langle w-w^{\ast},H^{k}(w-w^{\ast})\rangle=\frac{1}{2}\|w-w^{\ast}\|_{H^{k}}^{2}
\end{align*}
\end{comment}
\begin{comment}
Let $\xi_{t}^{k}$ be the stochastic sample on the $k$th device at
time $t$, and define $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$
as the stochastic Hessian matrix. %	
\end{comment}
\begin{comment}
Note that $\mathbb{E}\tilde{H}_{t}^{k}=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}=H^{k}$
and $\mathbf{g}_{t,k}=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$
while $\mathbf{g}_{t}=\sum_{k=1}^{N}p_{k}\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\sum_{k=1}^{N}p_{k}\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$.
\end{comment}
Define $l$ to be the smallest positive number such that $\mathbb{E}\|\xi_{t}^{k}\|^{2}$$\mathbf{\xi}_{t}^{k}$($\mathbf{\xi}_{t}^{k})^{T}\preceq lH^{k}$
for all $k$. Note that $l\leq\max_{k,j}\|\mathbf{x}_{k}^{j}\|^{2}$.
Let $L$ and $\mu$ be lower and upper bounds of non-zero eigenvalues
of $H^{k}$. Define $\kappa_{1}:=l/\mu$ and $\kappa:=L/\mu$. Following
\cite{liu2018accelerating,jain2017accelerating}, we define the statistical
condition number $\tilde{\kappa}$ as the smallest positive real number
such that $\mathbb{E}\left[\langle\xi_{t}^{k}(H^{k})^{-1},\xi_{t}^{k}\rangle\xi_{t}^{k}(\xi_{t}^{k})^{T}\right] \  \preceq\tilde{\kappa}H^{k}$, for all $k$. 
The condition numbers $\kappa_{1}$ and $\tilde{\kappa}$
are important in the characterization of convergence rates for FedAvg
algorithms. Note that $\kappa_{1}>\kappa$ and $\kappa_{1}>\tilde{\kappa}$.

@@

We show that the bound in Theorem \ref{thm:overparameterized_general} can be improved to $\mathcal{O}(\exp(-\frac{N}{E\kappa_{1}}t))$ for quadratic problems, then propose a variant of FedAvg that has provable acceleration over FedAvg with SGD updates.

>>>>

@@

We now turn to quadratic problems and show that the bound in Theorem \ref{thm:overparameterized_general} can be improved to $\mathcal{O}(\exp(-\frac{N}{E\kappa_{1}}t))$. We then propose a variant of FedAvg that has provable acceleration over FedAvg with SGD updates. The local device objectives are now given by the sum of squares {\small$F_{k}(\mathbf{w})=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(\mathbf{w}^{T}\mathbf{x}_{k}^{j}-z_{k}^{j})^{2}$},
and there exists $\mathbf{w}^{\ast}$ such that $F(\mathbf{w}^{\ast})\equiv0$. Two notions of condition number are important in our results:$\kappa_1$ which is based on local Hessians, and $\tilde{\kappa}$, which is termed the statistical condition number~\cite{liu2018accelerating,jain2017accelerating}. For their detailed definitions, please refer to Appendix Section~\ref{sec:interpolation}. Here we use the fact $\tilde{\kappa} \leq \kappa_1$.

\begin{comment}
Define the local Hessian matrix as $H^{k}:=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}$, and the stochastic Hessian matrix as $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$. Define $l$ to be the smallest positive number such that $\mathbb{E}\|\xi_{t}^{k}\|^{2}$$\mathbf{\xi}_{t}^{k}$($\mathbf{\xi}_{t}^{k})^{T}\preceq lH^{k}$
for all $k$. Note that $l\leq\max_{k,j}\|\mathbf{x}_{k}^{j}\|^{2}$.
Let $L$ and $\mu$ be lower and upper bounds of non-zero eigenvalues
of $H^{k}$. Define $\kappa_{1}:=l/\mu$ and $\kappa:=L/\mu$. Following
\cite{liu2018accelerating,jain2017accelerating}, we define the statistical
condition number $\tilde{\kappa}$ as the smallest positive real number
such that $\mathbb{E}\left[\langle\xi_{t}^{k}(H^{k})^{-1},\xi_{t}^{k}\rangle\xi_{t}^{k}(\xi_{t}^{k})^{T}\right] \  \preceq\tilde{\kappa}H^{k}$, for all $k$. 
The condition numbers $\kappa_{1}$ and $\tilde{\kappa}$
are important in the characterization of convergence rates for FedAvg
algorithms. Note that $\kappa_{1}>\kappa$ and $\kappa_{1}>\tilde{\kappa}$.
\end{comment}
%details to be moved to the appendix.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{comment}
In general $H$ has zero eigevalues. However, because the null space
of $H$ and range of $H$ are orthogonal, in our subsequence analysis
it suffices to project $\overline{\mathbf{w}}_{t}-\mathbf{w}^{\ast}$
onto the range of $H$, thus we may restrict to the non-zero eigenvalue
of $H$. We can use $\mathbf{w}^{\ast T}\mathbf{x}_{k}^{j}-z_{k}^{j}\equiv0$
to rewrite the local objectives as $F_{k}(\mathbf{w})=\frac{1}{2}\langle\mathbf{w}-\mathbf{w}^{\ast},H^{k}(\mathbf{w}-\mathbf{w}^{\ast})\rangle\equiv\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H^{k}}^{2}$
so that $F(\mathbf{w})=\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H}^{2}$.
\end{comment}
\begin{comment}
\begin{align*}
F_{k}(w) & =\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(w^{T}x_{k,j}-z_{k,j}-(w^{\ast T}x_{k,j}-z_{k,j}))^{2}=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}((w-w^{\ast})^{T}x_{k,j})^{2}\\
& =\frac{1}{2}\langle w-w^{\ast},H^{k}(w-w^{\ast})\rangle=\frac{1}{2}\|w-w^{\ast}\|_{H^{k}}^{2}
\end{align*}
\end{comment}
\begin{comment}
Let $\xi_{t}^{k}$ be the stochastic sample on the $k$th device at
time $t$, and define $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$
as the stochastic Hessian matrix. %	
\end{comment}
\begin{comment}
Note that $\mathbb{E}\tilde{H}_{t}^{k}=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}=H^{k}$
and $\mathbf{g}_{t,k}=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$
while $\mathbf{g}_{t}=\sum_{k=1}^{N}p_{k}\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\sum_{k=1}^{N}p_{k}\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$.
\end{comment}
